
\author{Susan B. Anthony}
\authdate{1820--1906}
\textdate{1873}
\chapter{Speech on Receiving Her Sentence}
\source{dailydemocrat1874}

\page{81}\noindent \textsc{Judge Hunt}---(Ordering the defendant to
stand up), ``Has the prisoner anything to say why sentence shall not
be pronounced?''

\page{82}\textsc{Miss Anthony}---Yes, your honor, I have many things
to say; for in your ordered verdict of guilty, you have trampled under
foot every vital principle of our government. My natural rights, my
civil rights, my political rights, my judicial rights, are all alike
ignored. Robbed of the fundamental privilege of citizenship, I am
degraded from the status of a citizen to that of a subject; and not
only myself individually, but all of my sex, are, by your honor's
verdict, doomed to political subjection under this, so-called, form of
government.

\textsc{Judge Hunt}---The Court cannot listen to a rehearsal of
arguments the prisoner's counsel has already consumed three hours in
presenting.

Miss Anthony---May it please your honor, I am not arguing the
question, but simply stating the reasons why sentence cannot, in
justice, be pronounced against me. Your denial of my citizen's right
to vote, is the denial of my right of consent as one of the governed,
the denial of my right of representation as one of the taxed, the
denial of my right to a trial by a jury of my peers as an offender
against law, therefore, the denial of my sacred rights to life,
liberty, property and---

\textsc{Judge Hunt}---The Court cannot allow the prisoner to go on.

\textsc{Miss Anthony}---But your honor will not deny me this one and
only poor privilege of protest against this high-handed outrage upon
my citizen's rights. May it please the Court to remember that since
the day of my arrest last November, this is the first time that
either myself or any person of my disfranchised class has been allowed
a word of defense before judge or jury---

\textsc{Judge Hunt}---The prisoner must sit down---the Court cannot
allow it.

\page{83}\textsc{Miss Anthony}---All of my prosecutors, from the 8th
ward corner grocery politician, who entered the complaint, to the
United States Marshal, Commissioner, District Attorney, District
Judge, your honor on the bench, not one is my peer, but each and all
are my political sovereigns; and had your honor submitted my case to
the jury, as was clearly your duty, even then I should have had just
cause of protest, for not one of those men was my peer; but, native or
foreign born, white or black, rich or poor, educated or ignorant,
awake or asleep, sober or drunk, each and every man of them was my
political superior; hence, in no sense, my peer. Even, under such
circumstances, a commoner of England, tried before a jury of Lords,
would have far less cause to complain than should I, a woman, tried
before a jury of men. Even my counsel, the Hon. Henry R. Selden, who
has argued my cause so ably, so earnestly, so unanswerably before your
honor, is my political sovereign. Precisely as no disfranchised person
is entitled to sit upon a jury, and no woman is entitled to the
franchise, so, none but a regularly admitted lawyer is allowed to
practice in the courts, and no woman can gain admission to the
bar---hence, jury, judge, counsel, must all be of the superior class.

\textsc{Judge Hunt}---The Court must insist---the prisoner has been
tried according to the established forms of law.

\textsc{Miss Anthony}---Yes, your honor, but by forms of law all made
by men, interpreted by men, administered by men, in favor of men, and
against women; and hence, your honor's ordered verdict of guilty,
against a United States citizen for the exercise of ``\textit{that
citizen's right to vote},'' simply because that citizen was a woman
and not a man. But, yesterday, the same man made forms of law,
declared it a crime punishable with \$1,000 fine and six months'
imprisonment, for you, or me, or any of us, to give a cup of cold
water, a crust \page{84} of bread, or a night's shelter to a panting
fugitive as he was tracking his way to Canada. And every man or woman
in whose veins coursed a drop of human sympathy violated that wicked
law, reckless of consequences, and was justified in so doing. As
then, the slaves who got their freedom must take it over, or under, or
through the unjust forms of law, precisely so, now, must women, to get
their right to a voice in this government, take it; and I have taken
mine, and mean to take it at every possible opportunity.

\textsc{Judge Hunt}---The Court orders the prisoner to sit down. It
will not allow another word.

\textsc{Miss Anthony}---When I was brought before your honor for
trial, I hoped for a broad and liberal interpretation of the
Constitution and its recent amendments, that should declare all United
States citizens under its protecting {\ae}gis---that should declare
equality of rights the national guarantee to all persons born or
naturalized in the United States. But failing to get this
justice---failing, even, to get a trial by a jury \textit{not} of my
peers---I ask not leniency at your hands---but rather the full rigors
of the law.

\textsc{Judge Hunt}---The Court must insist---

(Here the prisoner sat down.)

\textsc{Judge Hunt}---The prisoner will stand up.

(Here Miss Anthony arose again.)

The sentence of the Court is that you pay a fine of one hundred
dollars and the costs of the prosecution.

\textsc{Miss Anthony}---May it please your honor, I shall never pay a
dollar of your unjust penalty. All the stock in trade I possess is a
\$10,000 debt, incurred by publishing my paper---\textit{The
Revolution}---four years ago, the sole object of which was to educate
all women to \page{85} do precisely as I have done, rebel against your
man-made, unjust, unconstitutional forms of law, that tax, fine,
imprison and hang women, while they deny them the right of
representation in the government; and I shall work on with might and
main to pay every dollar of that honest debt, but not a penny shall go
to this unjust claim. And I shall earnestly and persistently continue
to urge all women to the practical recognition of the old
revolutionary maxim, that ``Resistance to tyranny is obedience to
God.''

\textsc{Judge Hunt}---Madam, the Court will not order you committed
until the fine is paid.

