
\author{Mary Astell}
\authdate{1666--1731}
\textdate{1700/1706}
\chapter[Mary Astell -- Some Reflections upon Marriage, excerpt]{Some
Reflections upon Marriage}

% The book was first published in 1700. The third edition of the book,
% published in 1706, contains the appendix, which is a longer version
% of what had been the book's preface. I have no idea whether Astell
% altered the text for the fourth edition.

\nfootnote{\fullcite{astell1730}}

%\page{63}Indeed Subjection, according to the common Notion of it, is
%not over easy; none of us, whether Men or Women, but have so good an
%Opinion of our own Conduct, as to believe we are fit, if not to direct
%others, at least to govern our selves. Nothing but a sound
%Understanding, and Grace, the best Improver of Natural Reason, can
%correct this Opi-\page{64}nion, truly humble us, and heartily
%reconcile us to Obedience. This better Cup therefore ought to be
%sweetned as much as my be; for Authority may be preserv'd and
%Government kept inviolable, without that nauseous Ostentation of
%Power, which serves to no End or Purpose, but to blow up the Pride and
%Vanity of those who have it, and to exasperate the Spirits of such as
%must truckle under it.

\page{64}Insolence is never the Effect of Power but in weak and
cowardly Spirits, who wanting true \textit{Merit} and Judgment to
support themselves in that Advantageous Ground on which they stand,
are ever appealing to their Authority, and making a Shew of it to
maintain their Vanity and Pride. A truly great Mind, and such as is
fit to Govern, tho' it may stand on its Right with its Equals, and
modestly expect what is due to it ever from its Superiors, yet it
never contends with its Inferiors, nor make use of its Superiority but
to do them Good. \page{65} So that considering the just Dignity of
Man, his great Wisdom so conspicuous on all Occasions! the Goodness
of his Temper, and Reasonableness of all his Commands, which make it a
Woman's Interest as well as Duty to be observant and obedience in all
Things; that his Prerogative is settled by an undoubted Right and the
Prescription of many Ages; it cannot be suppos'd, that he should
make frequent and insolent Claims of an Authority so well establish'd
and us'd with such Moderation, nor give an impartial By-stander (could
such an one be found) any Occasion from thence to suspect that he is
inwardly conscious of the Badness of his Title; Usurpers being
always most desirous of Recognitions, and busy in imposing Oaths,
whereas a Lawful Prince contents himself with the usual Methods and
Securities.

And since Power does naturally puff up, and he who finds himself
exalted, seldom fails to think he \textit{ought} to be so, it is more
suitable to a Man's Wisdom \page{66} and Generosity, to be mindful of
his great Obligations, than to insist on his Rights and Prerogatives.
Sweetness of Temper and an obliging Carriage are so justly due to a
Wife, that a Husband who must not be thought to want either
Understanding to know what is fit, nor Goodness to perform it, can't
be suppos'd not to shew them. For setting aside the Hazard of her
Person to keep up his Name and Family, with all the Pains and Trouble
that attend it, which may well be thought great enough to deserve all
the Respect and Kindness that may be; setting this aside, though 'tis
very considerable, a Woman has so much the Disadvantage in
\textit{most}, I was about to say, in \textit{all} Things, that she
makes a Man the greatest Compliment in the World when she condescends
to make him \textit{for Better for Worse}. She puts her self intirely
in his Power, leaves all that is dear to her, her Friends and Family,
to espouse his Interests and follow his Fortune, and makes it her
Business and Duty to please him! What Acknowledgements, \page{67} what
Returns can he make? What Gratitude can be sufficient for such
Obligations? She shews her good Opinion of him by the great Trust she
reposes in him, and what a Brute must he be who betrays that Trust, or
acts any way unworthy of it? Ingratitude is one of the basest Vices,
and if a Man's Soul is sunk so low as to be guilty of it towards her
who has so generously oblig'd him, and who so intirely depends on him,
if he can treat her discrepectfully, who has so fully testified her
Esteem of him, she must have a Stock of Vertue which he should blush
to discern, if she can pay him that Obedience of which he is so
unworthy.

Superiors indeed are too apt to forget the common Privileges of
Mankind; that their Inferiors share with them the greatest Benefits,
and are as capable as themselves of enjoying the supreme Good; that
though the Order of the World requires an \textit{Outward} Respect and
Obedience from some to others, yet the Mind \page{68} is free, nothing
but Reason can oblige it, 'tis out of the Reach of the most absolute
Tyrant. Nor will it ever be well either with those who Rule or those
in Subjection, even from the Throne to every private Family, till
those in Authority look on themselves as plac'd in that Station for
the Good and Improvement of their Subjects, and not for their own
Sakes; not as the Reward of their Merit, or that they may prosecute
their own Desires and fulfil all their Pleasure, but as the
Representatives of \textsc{God}, whom they ought to imitate in the
Justice and Equity of their Laws, in doing Good and communicating
Blessings to all beneath them: By which, and not by following the
imperious Dictates of their own Will, they become truly Great and
Illustrious, and worthily fill their Place. And the Governed for their
Part, ceasing to envy the Pomp and Name of Authority, should respect
their Governors as placed in \textsc{God's} stead, and contribute what
they can to ease them of their real Cares, by a chearful and ready
Com-\page{69}pliance, with their good Endeavours, and by affording
them the Pleasure of Success in which noble and generous Designs.

For, upon a due Estimate, Things are pretty equally divided; those in
Subjection, as they have a less Glorious, so they have an easier Task
and a less Account to give; Whereas he who Commands, has in a great
measure the Faults of others to answer for as well as his own. 'Tis
true, he has the Pleasure of doing more Good than a private Person
can, and shall receive the Reward of it when Time shall be no more, in
Compensation for the Hazards he runs, the Difficulties he at present
encounters, and the large Account he is to make hereafter. Which
Pleasure and Reward are highly desirable, and most worthy our Pursuit;
but they are Motives which such as Usurp on their Governors, and make
them uneasy in the due Discharge of their Duty, never propose. As for
those other little Things that movetheir Envy and Am-\page{70}bition,
they are of no Esteem with a just Considerer, nor will such as
violently pursue, find their Account in them.

But how can a Man respect his Wife when he has a contemptible Opinion
of her and her Sex? When from his own Elevation he looks down on them
as void of Understanding, full of Ignorance and Passion, so that Folly
and a Woman are equivalent Terms with him? Can he think there is any
Gratitude due to her whose utmost Services he exacts as strict Duty?
Becaus she was made to be a Slave to his Will, and has no higher End
than to Serve and Obey him? Perhaps we arrogate too much to our
selves, when we say this Material World was made for our Sakes: That
its Glorious Maker has given us the Use of it is certain; but when we
suppose any Thing to be made purely for our Sakes, because we have
Dominion over it, we draw a false Conclusion. As he should say the
People were made for the Prince who is set over them, \page{71} would
be thought to be out of his Senses as well as his Politicks. Yet even
allowing that \textsc{God}, who made every Thing in Number, Weight and
Measure, who never acts but for some great and glorious End, an End
agreeable to His Majesty; allowing that He created such a Number of
Rational Spirits merely to serve their Fellow Creatures, yet how are
these Lords and Masters help'd by the Contempt they shew of their poor
humble Vassals? Is it not rather an Hindrance to that Service they
expect, as being an undeniable and constant Proof how unworthy they
are to receive it?

None of God's Creatures, absolutely consider'd, are in their own
Nature contemptible; the meanest Fly, the poorest Insect has its Use
and Vertue. Contempt is scarce a Human Passion, one may venture to say
it was not in innocent Man, for till Sin came into the World, there
was nothing in it to be contemn'd. But Pride, which makes every Thing
serve its Purposes, wrested this Passion from \page{72} its only Use,
so that instead of being an Antidote against Sin, it is become a grand
Promoter of it, nothing making us more worthy of that Contempt we
shew, than when, poor, weak, dependent Creatures as we are! we look
down with Scorn and Disdain on others.

There is not a surer Sign of a noble Mind, a Mind very far advanc'd
towards Perfection, than the being able to bear Contempt and an unjust
Treatment from one's Superiors evenly and patiently. For inward Worth
and real Excellency are the true Ground of Superiority, and one Person
is not in reality better than another, but as he is more Wise and
Good. But this World being a Place of Trial, and govern'd by general
Laws, just Retributions being reserv'd for hereafter, Respect and
Obedience many times become due for Order's sake, to those who don't
otherwise deserve them. Now tho' Humility keeps us from over-valuing
our selves or viewing our Merit through a false and
mag-\page{73}nifying \textit{Medium}, yet it does not put out our
Eyes, it does not, it ought not to deprive us of that pleasing
Sentiment which attends our Acting as we ought to Act, which is, as it
were, a Foretaste of Heaven, our present Reward for doing what is just
and fit. And when a Superior does a mean and unjust thing, as all
Contempt of one's Neighbour is, and yet this does not provoke his
Inferiors to refuse that Observance which their Stations in the World
require, they cannot but have an inward Sense of their own real
Superiority, the other having no Pretence to it, at the same Time that
they pay him an outward Respect and Deference, which is such a
flagrant Testimony of the sincerest Love of Order, as proves their
Souls to be of the highest and noblest Rank.

A Man therefore for his own sake, and to give Evidence that he has a
Right to those Prerogatives he assumes, should treat Women with a
little more Humanity and Regard than is usually paid \page{74} them.
Your whisling Wits may scoff at them, and what then? It matters not,
for they rally every Thing though ever so sacred, and rail at the
Women commonly in very good Company. Religion, its Priests, and those
its most constant and regular Professors, are the usual Subjects of
their manly, mannerly and surprizing Jests. Surprizing indeed! not for
the Newness of the Thought, the Brightness of the Fancy, or Nobleness
of Expression, but for the good Assurance with which such Thread-bare
Jests are again and again repeated. But that your grave Dons, your
learned Men, and, which is more, your Men of Sense, as they would be
thought, should stoop so low as to make Invectives against the Women,
forget themselves so much as to jest with their Slaves, who have
enither Liberty, nor Ingenuity to make Reprizals; that they should
waste their Time, and debase their good Sense, which fits them for the
most weighty Affairs, such as are suitable to their profound Wisdom and
exalted Understand-\page{75}ings! to render those poor Wretches more
ridiculous and odious who are already in their Opinion sufficiently
contemptible, and find no better Exercise of their Wit and Satire,
than such as are not worth their Pains, though it were possible to
Reform them, this, this indeed may justly be wondred at!

I know not whether or no Women are allow'd to have Souls; if they
have, perhaps it is not prudent to provoke them too much, lest, silly
as they are, they at last recriminate, and then what polite and
well-bred Gentleman, though himself is concern'd, can forbear taking
that lawful Pleasure, which all who understand Raillery must taste,
when they find his Jests who insolently began to peck at his
Neighbour, return'd with Interest upon his own Head? And indeed Men
are too Humane, too Wise, to venture at it, did they not hope for this
Effect, and expect the Pleasure of finding their Wit turn to such
Account: For if it be lawful to pry into a Secret, \page{76} this is,
without doubt, the whole Design of those find Discourses which have
been made against the Women from our great Fore-Fathers to this
present Time! Generous Man has too much Bravery, he is too Just and
too Good to assault a defenceless Enemy, and if he did inviegh against
the Women, it was only to do them Service! For since neither his Care
of their Education, his hearty Endeavours to improve their Minds, his
wholesome Precepts, nor great Example could do them good, as his last
and kindest Essay, he resolv'd to try what Contempt would do, and
chose rather to expose himself by a seeming Want of Justice, Equity,
Ingenuity and Good-nature, than suffer Women to remain such vain and
insignificant Creatures as they have hitherto been reckon'd; and
truly, Women are some Degrees beneath what I have thus far thought
them, if they do not make the best Use of his Kindness, improve
themselves, and, like Christians, return it.

\begin{center}* * *\end{center}

\page{113}To wind up this Matter; If a Woman were duly principled, and
taught to know the World, especially the true Sentiments that Men have
of her, and the Traps they lay for her under so many gilded
Compliments, and such a \page{114} seemingly great Respect, that
Disgrace would be prevented which is brought upon too many Families;
Women would Marry more discreetly, and demean themselves better in a
married State, than some People say they do. The Foundation, indeed,
ought to be laid deep and strong, she should be made a good Christian,
and understand why she is so, and then she will be every thing else
that is Good. Men need keep no Spies on a Woman's Conduct, need have
no Fear of her Vertue, or so much as of her Prudence and Caution, were
but a due Sense of true Honour and Vertue awaken'd in her; were her
Reason excited and prepared to consider the Sophistry of those
Temptations which would persuade her from her Duty; and were she put
in a way to know that it is both her Wisdom and Interest to observe
it: she would then and duly examine and weigh all the Circumstances,
the Good and Evil of a married State, and not be surprized with
unforeseen Inconveniencies, and either never consent to be a Wife, or
\page{115} make a good one when she does. This would shew her what
Human Nature \textit{is}, as well as what it \textit{ought} to be, and
teach her not only what she may justly expect, but what she must be
content with; would enable her to cure some Faults, and patiently to
suffer what she cannot cure.

Indeed nothing can assure Obedience, and render it what it ought to
be, but the Conscience of Duty, the paying it for \textsc{God's} sake.
Superiors don't rightly understand their own Interest when they
attempt to put out their Subjects Eyes to keep them Obedient. A blind
Obedience is what a Rational Creature should never pay, nor would such
an one receive it, did he rightly understand its Nature. For Human
Actions are no otherwise valuable, than as they are conformable to
Reason; but a blind Obedience is an Obeying \textit{without Reason},
for ought we know, \textit{against it}. \textsc{God} himself does not
require our Obedience at this rate; he lays before us the
Good-\page{116}ness and Reasonableness of his Laws, and were there any
thing in them whose Equity we could not readily comprehend, yet we
have this clear and sufficient Reason, on which, to found our
Obedience, that nothing but what's just and fit, can be enjoin'd by a
Just, a Wise, and Gracious \textsc{God}; but this is a Reason will
never hold in respect of Mens Commands, unless they can prove
themselves Infallible, and consequently Impeccable too.

It is therefore very much a Man's Interest, that Women should be good
Christians; for in this, as i every other Instance, he who does his
Duty, finds his own Account in it. Duty and true Interest are one and
the same Thing, and he who thinks otherwise is to be pitied for being
so much in the Wrong: But what can be more the Duty of the Head, than
to instruct and improve those who are under Government? She will
freely leave him the quiet Dominion of this World, whose Thoughts and
Expecta-\page{117}tions are plac'd on the next. A Prospect of Heaven,
and that only, will cure that Ambition which all generous Minds are
fill'd with, not by taking it away, but by placing it on a right
Object. She will discern a Time when her Sex shall be no Bar to the
best Employments, the highest Honour; a Time when that Distinction,
now so much us'd to her Prejudice, shall be no more; but, provided she
is not wanting to her self, her Soul shall shine as bright as the
greatest Heroe's. This is a true, and indeed, the only Consolation;
this makes her a sufficient Compensation for all the Neglect and
Contempt the ill-grounded Customs of the World throw on her; for all
the Injuries brutal Power may do her, and is a sufficient Cordial to
support her Spirits, be her Lot in this World what it may.

But some sage Persons may, perhaps object, that were Women allow'd to
improve themselves, and not, amongst other Discouragements, driven
back by the \page{118} wise Jests and Scoffs that are put upon a Woman
of Sense or Learning, a Philosophical Lady, as she is call'd by way of
Ridicule; they would be too wise, and too good for the Men: I grant
it, for vicious and foolish Men. Nor is it to be wonder'd that He is
afraid he should not be able to Govern them were their Understandings
improv'd, who is resolv'd not to take too much Pains with his own. But
these, 'tis to be hoped, are no very considerable Number, the Foolish
at least; and therefore this is so far from being an Argument against
Womens Improvement, that it is a strong one for it, if we do but
suppose the Men to be as capable of Improvement as the Women; but much
more, if, according to Tradition, we believe they have greater
Capacities. This, if any thing, would stire them up to be what they
ought, and not permit them to waste their Time and abuse their
Faculties in the Service of their irregular Appetites and unreasonable
Desires, and so let poor contemptible Women, who have \page{119} been
their Slaves, excel them in all that is truly excellent. This would
make them Blush at employing an immortal Mind no better than in making
Provision for the Flesh to fulfil the Lusts thereof, since Women, by a
wiser Conduct, have brought themselves to such a Reach of Thought, to
such Exactness of Judgment, such Clearness and Strength of Reasoning,
such Purity and Elevation of Mind, such Command of their Passions,
such Regularity of Will and Affection, and, in a Word, to such a Pitch
of Perfection, as the Human Soul is capable of attaining in this Life
by the Grace of \textsc{God}; such true Wisdom, such real Greatness,
as though it does not qualify them to make a Noise in this World, to
found or overturn Empires, yet it qualifies them for what is
infinitely better, a Kingdom that cannot be mov'd, an incorruptible
Crown of Glory.

Besides, it were ridiculous to suppose, that a Woman, were she ever so
much improv'd, could come near the \page{120} topping Genius of the
Men, and therefore why should they envy or discourage her? Strength of
Mind goes along with Strength of Body, and 'tis only for some odd
Accidents which Philosophers have not yet thought worth while to
enquire into, that the sturdiest Porter is not the wisest Man! As
therefore the Men have the Power in their Hands, so there's no Dispute
of their having the Brains to manage it! Can we suppose there is such
a Thing as good Judgment and Sense upon Earth, if it is not to be
found among them: Do not they, generally speaking, do all the great
Actions and considerable Business of this World, and leave that of the
next to the Women? Their Subtlety in forming Cabals and laying deep
Designs, their Courage and Conduct in breaking through all Tyes,
sacred and civil, to effect them, not only advances them to the Post
of Honour, and keeps them securely in it for tenty or thirty Years,
but gets them a Name, and conveys it down to Posterity for some
Hundreds; and who would look \page{121} any further? Justice and
Injustice are administred by their Hands, Courts and Schools are
fill'd with these Sages; 'tis Men who dispute for Truth, as well as
Men who argue against it: Histories are writ by them; they recount
each other's great Exploits, and have always done so. All famous Arts
have their Original from Men, even from the Invention of Guns, to the
Mystery of good Eating. And to shew that nothing is beneath their
Care, any more than above their Reach, they have brought
\textit{Gaming} to an Art and Science, and a more Profitable and
Honourable one too, than any of those that us'd to be call'd
\textit{Liberal!} Indeed, what is they can't perform, when they
attempt it? The Strength of their Brains shall be every whit as
conspicuous at their Cups, as in a Senate-House, and, when they
please, they can make it pass for as sure a Mark of Wisdom, to drink
deep, as to reason profoundly; a greater Proof of Courage, and
consequently of Understanding, to dare the Vengeance of Heaven it
self, \page{122} than to stand the Raillery of some of the worst of
their Fellow Creatures!

Again, it may be said, If a Wife's Case be as it is here represented,
it is not good for a Woman to marry, and so there's an End of Human
Race. But this is no fair Consequence, for all that can justly be
inferr'd from hence, is, that a Woman has no mighty Obligations to the
Man who makes Love to her; she has no Reason to be fond of being a
Wife, or to reckon it a Piece of Preferment when she is taken to be a
Man's Upper-Servant; it is no Advantage to her in this World; if
rightly manag'd it may prove one as to the next. For she who marries
purely to do good, to educate Souls for Heaven, who can be so truly
mortified as to lay aside her own Will and Desires, to pay such an
intire Submission for Life, to one whom she cannot be sure will always
deserve it, does certainly perform a more Heroick Action, than all the
fa-\page{123}mous Masculine Heroes can boast of, she suffers a
continual Martyrdom to bring Glory to \textsc{God}, and Benefit to
Mankind; which Consideration, indeed, may carry her through all
Difficulties, I know not what else can, and engage her to Love him who
proves perhaps so much worse than a Brute, as to make this Condition
yet more grievous than it needed to be. She has need of a strong
Reason, of a truly Christian and well-termper'd Spirit, of all the
Assistance the best Education can give her, and ought to have some
good Assurance of her own Firmness and Vertue, who ventures on such a
Trial; and for this Reason 'tis less to be wonder'd at that Women
marry off in haste, for perhaps if they took Time to consdier and
reflect upon it, they seldom would marry.

\page{133}\section*{Appendix}

The \textit{Reflector}, who hopes \textit{Reflector} is not bad
\textit{English}, (now \textit{Governor} is happily of the Feminine
Gender) guarded against Curiousity in vain: For a certain ingenuous
Gentleman, as she is inform'd, had the Good-nature to own these
Reflections, so far, as to affirm he had the Original \textit{MS}. in
his Closet, a Proof she is not able to produce; and so to make himself
responsible for all their Faults, for which, she returns him all due
Acknowledgment. However, the Generality being of Opinion, that a Man
would have had more Prudence and Manners than to have Publish'd such
\page{134} unseasonable Truths, or to have betray'd the \textit{Arcana
Imperii} of his Sex; she humbly confesses, that the Contrivance and
Execution of this Design, which is unfortunately accus'd of being so
destructive to the Government, (of the Men, I mean) is intirely her
own. She neither advis'd with Friends, nor turn'd over antient or
modern Authors, nor prudently submitted to the Correction of such as
are, or such as \textit{think} they are good Judges, but with an
\textit{English} Spirit and Genius, set out upon the Forlorn Hope,
meaning no Hurt to any body, nor designing any thing but the publick
Good, and to retrieve, if possible, the Native Liberty, the Rights and
Privileges of the Subject.

Far be it from her to stir up Sedition of any sort: none can abhor it
more; and she heartily wishes, that our Masters would pay their Civil
and Ecclesiastical Governors the same Submission, wich they themselves
exact from their Do-\page{135}mestick Subjects. Nor can she imagine
how she any way undermines the Masculine Empire, or blows the Trumpet
of Rebellin to the Moiety of Mankind. Is it by exhorting Women, not to
expect to have their own Will in any thing, but to be intirely
Submissive, when once they have made Choice of a Lord and Master,
though he happen not to be so wise, so kind, or even so just a
Governor as was expected? She did not, indeed, advise them to think
his Folly Wisdom, nor his Brutality, that Love and Worship he promised
in his Matrimonial Oath; for this required a Flight of Wit and Sense
much above her poor Ability, and proper only to Masculine
Understandings. However, she did not in any manner prompt them to
Resist, or to Abdicate the Perjur'd Spouse, though the Laws of
\textsc{God}, and the Land, make special Provision for it, in a Case,
wherein, as is to be fear'd, few Men can truly plead Not Guilty.

\page{136}'Tis true, through want of Learning, and of that Superior
Genius which Men, as Men, lay claim to, she was ignorant of the
\textit{Natural Inferiority} of our Sex, which our Masters lay down as
a Self-evident and Fundamental Truth. She saw nothing in the Reason of
Things, to make this either a Principle or a Conclusion, but much to
the contrary; it being Sedition at least, if not Treason, to assert it
in this Reign. For if by the Natural Superiority of their Sex, they
mean, that \textit{every} Man is by Nature superior to \textit{every}
Woman, which is the obvious Meaning, and that which must be stuck to
if they would speak Sense, it would be a Sin in \textit{any} Woman, to
have Dominion over \textit{any} Man, and the greatest Queen ought not
to command, but to obey, her Footman: because no Municipal Laws can
supersede or change the Law of Nature: So that if the Dominion of the
Men be such, the \textit{Salique Law}, as unjust as \textit{English
Men} have ever thought it, ought to Place over all \page{137} the
Earth, and the most glorious Reigns in the \textit{English, Danish,
Castilian}, and other Annals, were wicked Violations of the Law of
Nature!

If they mean that \textit{some} Men are superior to \textit{some}
Women, this is no great Discovery; had they turn'd the Tables, they
might have seen that \textit{some} Women are superior to \textit{some}
Men. Or had they been pleased to remember their Oaths of Allegiance
and Supremacy, they might have known, that \textit{One} Woman is
superior to \textit{All} the Men in these Nations, or else they have
sworn to very little Purpose. And it must not be suppos'd, that their
Reason and Religion would suffer them to take Oaths, contrary to the
Law of Nature and Reason of Things.

\begin{center}* * *\end{center}

\page{144}Our \textit{Reflector} is of Opinion, that Disputes of this
kind, extending to Human Nature in general, and not peculiar to those
to whom the Word of \textsc{God} has been reveal'd, ought to be
decided by Natural Reason only. And, that the Holy Scripture should
not be interested in the present Controversy, in which it determines
nothing, any more than it does between the \textit{Copernican} and
\textit{Ptolomean} Systems. The Design of those Holy Books being to
make us excellent Moralists and perfect Christians, not great
Philosophers; and being writ for the Vulgar as well as for the
Learned, they are accommodated to the common way of Speech and the
Usage of the World; in which we have but a short Probation, so that it
matters not much what Part we act, whether of Governing or Obeying,
provided we perform it well with respect to the World to come.

One does not wonder, indeed, that when an Adversary is drove to a
Non-\page{145}plus, and Reason declares against him, he flies to
Authority, especially to Divine, which is infallible, and therefore
ought not to be disputed. But Scripture is not always on their Side
who make Parade of it, and through their Skill in Languages, and the
Tricks of the Schools, wrest it from its genuine Sense to their own
Inventions. And supposing, not granting, that it were apparently to
the Womens Disadvantage, no fair and generous Adversary but would be
asham'd to urge this Advantage: Because Women, without their own
Fault, are kept in Ignorance of the Original, wanting Languages and
other Helps to Criticise on the Sacred Text, of which, they know no
more, than Men are pleas'd to impart in their Translations. In short,
they shew their Desire to maintain their Hypotheses, but by no means
their Reverence to the Sacred Oracles, who engage them in such
Disputes. And therefore, the Blame be theirs, who have unnecessarily
introduc'd them in \page{146} the present Subject, and who, by saying,
that the \textit{Reflections} were not agreeable to Scripture, oblige
the Reflector to shew, that those who affirm it must either mistake
her Meaning, or the Sense of Holy Scripture, or both, if they think
what they say, and do not find fault meerly because they resolve to do
so. For, had she ever writ any thing contrary to those sacred Truths,
she would be the first in pronouncing its Condemnation.

But what says the Holy Scripture? It speaks of Women as in a State of
Subjection, and so it does of the \textit{Jews} and
\textit{Christians}, when under the Dominion of the \textit{Chaldeans}
and \textit{Romans}, requiring of the one as well as of the other, a
quiet Submission to them under whose Power they liv'd. But will any
one say, that these had a \textit{Natural Superority} and Right to
Dominion? that they had a superior Understanding, or any Pre-eminence,
except what their greater Strength acquir'd? Or, that the other were
sub-\page{147}jected to their Adversaries for any other Reason but the
Punishment of their Sins, and, in order to their Reformation? Or for
the Excercise of their Vertue, and because the Order of the World and
the Good of Society requir'd it?

If Mankind had never Sin'd, Reason would always have been obeyed,
there would have been no Struggle for Dominion, and Brutal Power would
not have prevail'd. But in the lapsed State of Mankind, and now, that
Men will not be guided by their Reason but by their Appetites, and do
not what they \textit{ought} but what they \textit{can}, the Reason,
or that which stands for it, the Will and Pleasure of the Governor, is
to be the Reason of those who will not be guided by their own, and
must take Place for Order's sake, although it should not be
conformable to right Reason. Nor can there be any Society great or
little, from Empires down to private Families, without a last Resort,
to determine the Affairs of \page{148} that Society by an irrestitable
Sentence. Now unless this Supremacy be fix'd somewhere, there will be
a perpetual Contention about it, such is the Love of Dominion, and let
the Reason of Things be what it may, those who have least Force or
Cunning to supply it, will have the Disadvantage. So that since Women
are acknowledged to have the least Bodily Strength, their being
commanded to Obey is in pure Kindness to them, and for their Quiet and
Security, as well as for the Exercise of their Vertue. But does it
follow, that Domestick Governors have more Sense than their Subjects,
any more than that other Governors have? We do not find that any Man
thinks the worse of his own Understanding, because another has
superior Power; or concludes himself less capable of a Post of Honour
and Authority, because he is not prefer'd to it. How much Time would
like on Mens Hands, how empty would the Places of Concourse be, and
how silent most Companies, did \page{149} Men forbear to censure their
Governors, that is, in effect, to think themselves wiser. Indeed,
Government would be much more desirable than it is, did it invest the
Possessor with a superior Understanding as well as Power. And if meer
Power gives a Right to Rule, there can be no such Thing as Usurpation;
but a Highway-Man, so long as he has Strength to force, has also a
Right to require our Obedience.

Again, if absolute Sovereignty be not necessary in a State, how comes
it to be so in a Family? Or if in a Family why not in a State; since
no Reason can be alleged for the one that will not hold more strongly
for the other? If the Authority of the Husband, so far as it extends,
is sacred and inalienable, why not that of the Prince? The Domestick
Sovereign is without Dispute elected, and the Stipulations and
Contract are mutual; is it not then partial in Men to the last Degree,
to contend for, and practise \page{150} that Arbitrary Dominion in
their Families, which they abhor and exclaim against in the State? For
if Arbitrary Power is evil in it self, and an improper Method of
Governing Rational and Free Agents, it ought not to be practis'd any
where; nor is it less, but rather more mischievous in Families than in
Kingdoms, by how much 100,000 Tyrants are worse than one. What though
a Husband can't deprive a Wife of Life without being responsible to
the Law, he may, however, do what is much more grievous to a generous
Mind, render Life miserable, for which she has no Redress, scarce
Pity, which is afforded to every other Complainant, it being thought a
Wife's Duty to suffer every thing without Complaint. If \textit{all
Men are born Free}, how is it that all Women are born Slaves? As they
must be, if the being subjected to the \textit{incostant, uncertain,
unknown, arbitrary Will} of Men, be the \textit{perfect Condition of
Slavery?} And, if the Essence of Freedom consists, as our
Ma-\page{151}sters say it does, in having a \textit{standing Rule to
live by?} And why is Slavery so much condemn'd and strove against in
one Case, and so highly applauded, and held so necessary and so sacred
in another?

\begin{center}* * *\end{center}

\page{171}However, there are strong and prevalent Reasons which
demonstrate the Superority and Pre-eminence of the Men. \page{172} For
in the first Place, Boys have much Time and Pains, Care and Cost
bestow'd on their Education, Girls have little or none. The former are
early initiated in the Sciences, are made acquainted with antient and
modern Discoveries, they study Books and Men, have all imaginable
Encouragement; not only Fame, a dry Reward now a-days, but also Title,
Authority, Power, and Riches themselves, which purchase all Things,
are the Reward of their Improvement. The latter are restrain'd,
frown'd upon, and beat, not \textit{for}, but \textit{from} the Muses;
Laughter and Ridicule, that never-failing Scare-Crow, is set up to
drive them from the Tree of Knowledge. But if, in spite of all
Difficulties Nature prevails, and they can't be kept so ignorant as
their Masters would have them, they are star'd upon as Monsters,
censur'd, envied, and every way discouraged, or, at the best, they
have the Fate the Proverb assigns them, \textit{Vertue is prais'd and
starv'd}. And therefore, since the coarsest \page{173} Materials need
the most Curing, as every Workman can inform you, and the worst Ground
the most elaborate Culture, it undeniably follows, that Mens
Understandings are superior to Womens, for, after many Years Study and
Experience, they become wise and learned, and Women are not Born so!

Again, Men are possessed of all Places of Power, Trust and Profit,
they make Laws and exercise the Magistracy, not only the sharpest
Sword, but even all the Swords and Blunderbusses are theirs, which by
the strongest Logick in the World, gives them the best Title to every
Thing they please to claim as their Prerogative: Who shall contend
with them? Immemorial Prescription is on their Side in these Parts of
the World, antient Tradition and modern Usage! Our Fathers, have all
along, both taught and practised Superiority over the weaker Sex, and
consequently Women are by Nature inferior to Men, as was to be
de-\page{174}monstrated. An Argument which must be acknowledged
unanswerable; for, as well as I love my Sex, I will not pretend a
Reply to \textit{such} Demonstration!

Only let me beg to be inform'd, to whom we poor Fatherless Maids, and
Widows who have lost their Masters, owe Subjection? It can't be to all
Men in general, unless all Men were agreed to give the same Commands;
Do we then fall as Strays, to the first who finds us? By the Maxims of
some Men, and the Conduct of some Women one would think so. But
whoever he be that thus happens to become our Master, if he allows us
to be reasonable Creatures, and doe snot meerly Compliment us with
that Title, since no Man denies our Readiness to use our Tongues, it
would tend, I should think, to our Master's Advantage, and therefore
he may please to be advis'd to teach us to improve our Reason. But if
Reason is only allow'd us by way of Raillery, and the secret
\page{175} Maxim is, that we have none, or little more than Brutes,
'tis the best way to confine us with Chain and Block to the
Chimney-Corner, which, probably, might save the Estates of some
Families and the Honour of others.

I do not propose this to prevent a Rebellion, for Women are not so
well united as to form an Insurrection. They are for the most part
wise enough to love their Chains, and to discern how very becomingly
they fit. They think as humbly of themselves as their Masters can
wish, whith repsect to the other Sex, but in regard to their own, they
have a Spice of Masculine Ambition; every one would Lead, and none
would Follow. Both Sexes being too apt to Envy, and too backward in
Emulating, and take more Delight in detracting from their Neighbour's
Vertue, than in improving their own. And therefore, as to those Women
who find themselves born for Slavery, and are so sensible of their own
\page{176} Meanness, as to conclude it impossible to attain to any
thing excellent, since they are, or ought to be best acquainted with
their own Strength and Genius, She's a Fool who would attempt their
Deliverance or Improvement. No, let them enjoy the great Honour and
Felicity of their tame, submissive and depending Temper! Let the Men
applaud, and let them glory in this wonderful Humility! Let them
receive the Flatteries and Grimaces of the other Sex, live unenvied by
their own, and be as much belov'd as one such Woman can afford to love
another! Let them enjoy the Glory of treading in the Footsteps of
their Predecessors, and of having the Producence to avoid that
audacious Attempt of soaring beyond their Sphere! Let them Huswife or
Play, Dress, and be pretty entertaining Company! Or, which is better,
relieve the Poor to ease their own Compassions, read pious Books, say
their Prayers, and go to Church, because they have been taught and
us'd to do so, \page{177} without being able to give a better Reason
for their Faith and Practice! Let them not by any means aspire at
being Women of Understanding, because no Man can endure a Woman of
Superior Sense, or would treat a reasonable Woman civilly, but that
the thinks he stands on higher Ground, and, that she is so wise as to
make Exceptions in his Favour, and to take her Measures by his
Directions; they may pretend to Sense, indeed, since meer Pretences
only render one the more ridiculous! Let them, in short, be what is
call'd \textit{very} Women, for this is most acceptable to all sorts
of Men; or let them aim at the Title of \textit{good devout} Women,
since some Men can bear with this; but let them not judge of the Sex
by their own Scantling: For the great Author of Nature and Fountain of
all Perfection, never design'd that the Mean and Imperfect, but that
the most Compleat and Excellent of His Creatures in every Kind, should
be the Standard of the rest.

