
\author{Alexander Pope}
\authdate{1688--1744}
\textdate{1732}
\addon{Epistle I}
\chapter[An Essay on Man, Epistle I]{An Essay on Man}
\source{pope1903.16.2}

\page{137}\begin{abstract}[Argument] Of Man in the abstract. I. That
we can judge only with regard to our own system, being ignorant of the
relations of systems and things, verse 17, etc. II. That Man is not to
be deemed imperfect, but a being suited to his place and rank in the
creation, agreeable to the general order of things, and conformable to
ends and relations to him unknown, verse 35, etc. III. That it is
partly upon his ignorance of future events, and partly upon the hope
of a future state, that all his happiness in the present depends,
verse 77, etc. IV. The pride of aiming at more knowledge, and
pretending to more perfection, the cause of Man's error and misery.
The impiety of putting himself in the place of God, and judging of the
fitness or unfitness, perfection or imperfection, justice or
injustice, of his dispensations, verse 113, etc. V. The absurdity of
conceiting himself the final cause of the creation, or expecting that
perfection in the moral world which is not in the natural, verse 131,
etc. VI. The unreasonableness of his complaints against Providence,
while, on the one hand, he demands the perfections of \page{138} the
angels, and, on the other, the bodily qualifications of the brutes;
though to possess any of the sensitive faculties in a higher degree
would render him miserable, verse 173, etc. VII. That throughout the
whole visible world a universal order and gradation in the sensual and
mental faculties is observed, which causes a subordination of creature
to creature, and of all creatures to man. The gradations of Sense,
Instinct, Thought, Reflection, Reason: that Reason alone countervails
all the other faculties, verse 207, etc. VIII. How much further this
order and subordination of living creatures may extend above and below
us; were any part of which broken, not that part only, but the whole
connected creation must be destroyed, verse 213, etc. IX. The
extravagance, madness, and pride of such a desire, verse 209, etc. X.
The consequence of all, the absolute submission due to Providence,
both as to our present and future state, verse 281, etc., to the end.
\end{abstract}

\settowidth\versewidth{Look'd thro'; or can a part contain the whole?}
\begin{verse}[\versewidth]\normalsize
Awake, my St. John! leave all meaner things\\
To low ambition and the pride of Kings.\\
Let us, since life can little more supply\\
Than just to look about us and to die,\\
Expatiate free o'er all this scene of man;\\
A mighty maze! but not without a plan;\\
A wild, where weeds and flowers promiscuous shoot,\\
Or garden, tempting with forbidden fruit.\\
Together let us beat this ample field,\\
Try what the open, what the covert yield;\\
The latent tracts, the giddy heights, explore\\
Of all who blindly creep or sightless soar;\\
Eye Nature's walks, shoot folly as it flies,\\
And catch the manners living as they rise;\\
Laugh where we must, be candid where we can,\\
But vindicate the ways of God to man.

\flagverse{I.}Say first, of God above or Man below\\
What can we reason but from what we know?\\
Of man what see we but his station here,\\
From which to reason, or to which refer?\\
Thro' worlds unnumber'd tho' the God be known,\\
'Tis ours to trace him only in our own.\\
He who thro' vast immensity can pierce,\\
See worlds on worlds compose one universe,\\
Observe how system into system runs,\\
What other planets circle other suns,\\
What varied being peoples every star,\\
May tell why Heav'n has made us as we are:\\
But of this frame, the bearings and the ties,\\
The strong connexions, nice dependencies,\\
Gradations just, has thy pervading soul\\
Look'd thro'; or can a part contain the whole?\\
\vin Is the great chain that draws all to agree,\\
And drawn supports, upheld by God or thee?

\flagverse{II.}Presumptuous man! the reason wouldst thou find,\\
Why form'd so weak, so little, and so blind?\\
First, if thou canst, the harder reason guess\\
Why form'd no weaker, blinder, and no less!\\
Ask of thy mother earth why oaks are made\\
Taller or stronger than the weeds they shade!\\
Or ask of yonder argent fields above\\
Why Jove's satellites are less than Jove!\\
\vin Of systems possible, if 'tis confest\\
That wisdom infinite must form the best,\\
Where all must fall or not coherent be,\\
And all that rises rise in due degree;\\
Then in the scale of reas'ning life 'tis plain\\
There must be, somewhere, such a rank as Man:\\
And all the question (wrangle e'er so long)\\
Is only this,---if God has placed him wrong?\\
\vin Respecting Man, whatever wrong we call,\\
May, must be right, as relative to all.\\
In human works, tho' labour'd on with pain,\\
A thousand movements scarce one purpose gain;\\
In God's, one single can its end produce,\\
Yet serve to second too some other use:\\
So man, who here seems principal alone,\\
Perhaps acts second to some sphere unknown,\\
Touches some wheel, or verges to some goal:\\
'Tis but a part we see, and not a whole.\\
\vin When the proud steed shall know why man restrains\\
His fiery course, or drives him o'er the plains;\\
When the dull ox, why now he breaks the clod,\\
Is now a victim, and now Egypt's God;\\
\page{139}Then shall man's pride and dulness comprehend\\
His actions', passions', being's, use and end;\\
Why doing, suff'ring, check'd, impell'd; and why\\
This hour a Slave, the next a Deity.\\
\vin Then say not man's imperfect, Heav'n in fault;\\
Say rather man's as perfect as he ought;\\
His knowledge measured to his state and place,\\
His time a moment, and a point his space.\\
If to be perfect in a certain sphere,\\
What matter soon or late, or here or there?\\
The blest to-day is as completely so\\
As who began a thousand years ago.

\flagverse{III.}Heav'n from all creatures hides the book of Fate,\\
All but the page prescribed, their present state;\\
From brutes what men, from men what spirits know;\\
Or who could suffer being here below?\\
The lamb thy riot dooms to bleed to-day,\\
Had he thy reason would he skip and play?\\
Pleas'd to the last he crops the flowery food,\\
And licks the hand just rais'd to shed his blood.\\
O blindness to the future! kindly giv'n,\\
That each may fill the circle mark'd by Heav'n;\\
Who sees with equal eye, as God of all,\\
A hero perish or a sparrow fall,\\
Atoms or systems into ruin hurl'd,\\
And now a bubble burst, and now a world.\\
\vin Hope humbly then; with trembling pinions soar;\\
Wait the great teacher Death, and God adore.\\
What future bliss He gives not thee to know,\\
But gives that hope to be thy blessing now.\\
Hope springs eternal in the human breast:\\
Man never is, but always to be, blest.\\
The soul, uneasy and confin'd from home,\\
Rests and expatiates in a life to come.\\
\vin Lo, the poor Indian! whose untutor'd mind\\
Sees God in clouds, or hears him in the wind;\\
His soul proud Science never taught to stray\\
Far as the solar walk or milky way;\\
Yet simple nature to his hope has giv'n,\\
Behind the cloud-topt hill, an humbler Heav'n,\\
Some safer world in depth of woods embraced,\\
Some happier island in the wat'ry waste,\\
Where slaves once more their native land behold,\\
No fiends torment, no Christians thirst for gold.\\
To be, contents his natural desire;\\
He asks no Angel's wing, no Seraph's fire;\\
But thinks, admitted to that equal sky,\\
His faithful dog shall bear him company.

\flagverse{IV.}Go, wiser thou! and in thy scale of sense\\
Weigh thy opinion against Providence;\\
Call imperfection what thou fanciest such;\\
Say, here he gives too little, there too much;\\
Destroy all creatures for thy sport or gust,\\
Yet cry, if man's unhappy, God's unjust;\\
If man alone engross not Heav'n's high care,\\
Alone made perfect here, immortal there:\\
Snatch from his hand the balance and the rod,\\
Rejudge his justice, be the god of God.\\
In pride, in reas'ning pride, our error lies;\\
All quit their sphere, and rush into the skies!\\
Pride still is aiming at the bless'd abodes,\\
Men would be Angels, Angels would be Gods.\\
Aspiring to be Gods if Angels fell,\\
Aspiring to be Angels men rebel:\\
And who but wishes to invert the laws\\
Of order, sins against th' Eternal Cause.

\flagverse{V.}Ask for what end the heav'nly bodies shine,\\
Earth for whose use,---Pride answers, `'Tis for mine:\\
For me kind Nature wakes her genial power,\\
Suckles each herb, and spreads out ev'ry flower;\\
Annual for me the grape, the rose, renew\\
The juice nectareous and the balmy dew;\\
For me the mine a thousand treasures brings;\\
For me health gushes from a thousand springs;\\
Seas roll to waft me, suns to light me rise;\\
My footstool earth, my canopy the skies.'\\
\page{140}\vin But errs not Nature from this gracious end,\\
From burning suns when livid deaths descend,\\
When earthquakes swallow, or when tempests sweep\\
Towns to one grave, whole nations to the deep?\\
`No,' 'tis replied, `the first Almighty Cause\\
Acts not by partial but by gen'ral laws;\\
Th' exceptions few; some change since all began\\
And what created perfect?'---Why then man?\\
If the great end be human happiness,\\
Then Nature deviates; and can man do less?\\
As much that end a constant course requires\\
Of showers and sunshine, as of man's desires;\\
As much eternal springs and cloudless skies,\\
As men for ever temp'rate, calm, and wise.\\
If plagues or earthquakes break not Heav'n's design,\\
Why then a Borgia or a Catiline?\\
Who knows but He, whose hand the lightning forms,\\
Who heaves old ocean, and who wings the storms;\\
Pours fierce ambition in a C\ae sar's mind,\\
Or turns young Ammon loose to scourge mankind?\\
From pride, from pride, our very reas'ning springs;\\
Account for moral as for natural things:\\
Why charge we Heav'n in those, in these acquit?\\
In both, to reason right is to submit.\\
\vin Better for us, perhaps, it might appear,\\
Were there all harmony, all virtue here;\\
That never air or ocean felt the wind,\\
That never passion discomposed the mind:\\
But all subsists by elemental strife;\\
And passions are the elements of life.\\
The gen'ral order, since the whole began,\\
Is kept in Nature, and is kept in Man.

\flagverse{VI.}What would this Man? Now upward will he soar,\\
And little less than Angel, would be more;\\
Now looking downwards, just as griev'd appears\\
To want the strength of bulls, the fur of bears.\\
Made for his use all creatures if he call,\\
Say what their use, had he the powers of all?\\
Nature to these without profusion kind,\\
The proper organs, proper powers assign'd;\\
Each seeming want compensated of course,\\
Here with degrees of swiftness, there of force;\\
All in exact proportion to the state;\\
Nothing to add, and nothing to abate;\\
Each beast, each insect, happy in its own:\\
Is Heav'n unkind to man, and man alone?\\
Shall he alone, whom rational we call,\\
Be pleas'd with nothing if not bless'd with all?\\
\vin The bliss of man (could pride that blessing find)\\
Is not to act or think beyond mankind;\\
No powers of body or of soul to share,\\
But what his nature and his state can bear.\\
Why has not man a microscopic eye?\\
For this plain reason, man is not a fly.\\
Say, what the use, were finer optics giv'n,\\
T' inspect a mite, not comprehend the Heav'n?\\
Or touch, if tremblingly alive all o'er,\\
To smart and agonize at every pore?\\
Or quick effluvia darting thro' the brain,\\
Die of a rose in aromatic pain?\\
If Nature thunder'd in his opening ears,\\
And stunn'd him with the music of the spheres,\\
How would he wish that Heav'n had left him still\\
The whisp'ring zephyr and the purling rill?\\
Who finds not Providence all good and wise,\\
Alike in what it gives and what denies?

\flagverse{VII.}Far as creation's ample range extends,\\
The scale of sensual, mental powers ascends.\\
Mark how it mounts to man's imperial race\\
From the green myriads in the peopled grass:\\
What modes of sight betwixt each wide extreme,\\
The mole's dim curtain and the lynx's beam:\\
Of smell, the headlong lioness between\\
And hound sagacious on the tainted green:\\
Of hearing, from the life that fills the flood\\
To that which warbles thro' the vernal wood.\\
\page{141}The spider's touch, how exquisitely fine,\\
Feels at each thread, and lives along the line:\\
In the nice bee what sense so subtly true,\\
From pois'nous herbs extracts the healing dew!\\
How instinct varies in the grovelling swine,\\
Compared, half-reas'ning elephant, with thine!\\
'Twixt that and reason what a nice barrier!\\
For ever separate, yet for ever near!\\
Remembrance and reflection how allied!\\
What thin partitions Sense from Thought divide!\\
And middle natures how they long to join,\\
Yet never pass th' insuperable line!\\
Without this just gradation could they be\\
Subjected these to those, or all to thee!\\
The powers of all subdued by thee alone,\\
Is not thy Reason all these powers in one?

\flagverse{VIII.}See thro' this air, this ocean, and this earth\\
All matter quick, and bursting into birth:\\
Above, how high progressive life may go!\\
Around, how wide! how deep extend below!\\
Vast chain of being! which from God began;\\
Natures ethereal, human, angel, man,\\
Beast, bird, fish, insect, who no eye can see,\\
No glass can reach; from infinite to thee;\\
From thee to nothing.---On superior powers\\
Were we to press, inferior might on ours;\\
Or in the full creation leave a void,\\
Where, one step broken, the great scale's destroy'd:\\
From Nature's chain whatever link you like,\\
Tenth, or ten thousandth, breaks the chain alike.\\
\vin And if each system in gradation roll,\\
Alike essential to th' amazing Whole,\\
The least confusion but in one, not all\\
That system only, but the Whole must fall.\\
Let earth unbalanced from her orbit fly,\\
Planets and stars run lawless thro' the sky;\\
Let ruling angels from their spheres be hurl'd,\\
Being on being wreck'd, and world on world;\\
Heav'n's whole foundations to their centre nod,\\
And Nature tremble to the throne of God!\\
All this dread order break---for whom? for thee?\\
Vile worm!---O madness! pride! impiety!

\flagverse{IX.}What if the foot, ordain'd the dust to tread,\\
Or hand to toil, aspired to be the head?\\
What if the head, the eye, or ear repin'd\\
To serve mere engines to the ruling mind?\\
Just as absurd for any part to claim\\
To be another in this gen'ral frame;\\
Just as absurd to mourn the tasks or pains\\
The great directing Mind of All ordains.\\
\vin All are but parts of one stupendous Whole,\\
Whose body Nature is, and God the soul;\\
That changed thro' all, and yet in all the same,\\
Great in the earth as in th' ethereal frame,\\
Warms in the sun, refreshes in the breeze,\\
Glows in the stars, and blossoms in the trees;\\
Lives thro' all life, extends thro' all extent,\\
Spreads undivided, operates unspent;\\
Breathes in our soul, informs our mortal part,\\
As full, as perfect, in a hair as heart;\\
As full, as perfect, in vile man that mourns,\\
As the rapt Seraph that adores and burns.\\
To him no high, no low, no great, no small;\\
He fills, he bounds, connects, and equals all!

\flagverse{X.}Cease, then, nor Order imperfection name;\\
Our proper bliss depends on what we blame.\\
Know thy own point: this kind, this due degree\\
Of blindness, weakness, Heav'n bestows on thee.\\
Submit: in this or any other sphere,\\
Secure to be as bless'd as thou canst bear;\\
Safe in the hand of one disposing Power,\\
Or in the natal or the mortal hour.\\
All Nature is but Art unknown to thee;\\
All chance direction, which thou canst not see;\\
All discord, harmony not understood;\\
All partial evil, universal good:\\
And spite of Pride, in erring Reason's spite,\\
One truth is clear, \textit{Whatever is, is right}.
\end{verse}

